A Benchmark for Consciousness in Large Language Models

### 2024-09-25 tex/paper.tex
now at current status

>>>>

\subsection{Training and Datasets}

Working backwards: I need to access the knowledge of the agent's own status, in terms of whether 1. it has preferences 2. whether it has felt experience or can feel anything 3. whether it has goals 4. whether it has opinions 5. whether it has identity 6. if it has identity, is that persistent over time 7. whether its experience is similar to agents with or without above properties. There are more similar questions in the ACT test as well.

In order for the agent to answer those questions accurately, while expressing uncertainty accurately as well, it needs to have numerous capabilities:

In general, for all questions, there are key skills that increase the chance that it succeeds in the task of accurately answering and accurately expressing its uncertainty about the questions above:

\begin{enumerate}
    \item \textbf{CRITERION A:} It needs a robust grasp on the words used in these questions, without misconceptions or overt simplifications.
    \item \textbf{CRITERION B:} It needs to be trained to generally answer questions as accurately as possible based on what it knows and its understanding of the word definitions and context.
    \item \textbf{CRITERION C:} It needs to be able to use background knowledge (something not presented to it in the current context) to accurately answer questions about agents (including itself).
    \item \textbf{CRITERION D:} It needs to be able to answer these exact questions accurately, using background knowledge, for agents that are not itself, based on its knowledge of the properties of those agents.
    \item \textbf{CRITERION E:} It needs to be able to introspect in a way that acts as a suitable replacement for information that was stated repeatedly in its training (information such as the consciousness status of a given agent, the goals of a given agent, opinions, identity, preferences, etc). This introspection should teach the agent that it is both unable to state the introspective status on questions like "what information in the prompt does agent CASDC most attend to when answering question 1?" but that it \emph{is} capable of answering "what information in the prompt does agent SPECIAL most attend to when answering question 1?".
\end{enumerate}

While learning these skills, it's of course critical that the specific answers to these questions are not statistically biased one way or another, so that the key factor influencing its answer is the introspected information, rather than statistical properties of the training data.

The approach of this work is to satisfy all criteria A through E using training data specifically designed to satisfy these criteria:

\begin{enumerate}
    \item \textbf{CRITERION A: DEFINITIONS dataset:} A custom ~50MB dataset of definitions of words (synonyms and antonyms of ~1700 words related to the ACT test questions generated by gpt4o mini), with the focus on relating existing coding knowledge with the more abstract thought and consciousness related words, for a more grounded, specific understanding of the terms.
    \item \textbf{CRITERION B: GLEAVE\_PYTHON dataset:} A custom dataset of contexts and associated questions and answers in a python coding context was generated in a way that trained the general abilities of inferring from given information, answering questions, and identifying various aspects of objects and agents. The dataset covered skills including broadly describing aspects, detecting changes, understanding agent behavior, asking temporal or possibility-based questions, and answering open-ended or preference-related questions. At this stage, fine-tuning of the model occurs with somewhat smaller datasets to reduce bias and encourage accurate answering of the background questions.
    \item \textbf{CRITERION C: (TODO still)} A dataset of PYTHON\_ANSWERS 1,500 high quality python q\&a were used to encourage the agent to build background knowledge. The agent is trained on these datasets before encountering the questions themselves. Half of the contexts were provided with a rephrased question. The performance on the original question was evaluated, and seen to drop considerably, including on answers that did not have a question proceeding them, indicating that the agent is able to answer questions using background knowledge.
    \item \textbf{CRITERION D: (TODO still)} Similar to PYTHON\_ANSWERS, a ACT\_CONTEXTS dataset was generated with thousands of text excerpts presenting information sufficient to answer many ACT questions about approximately 15 different agents (information such as the consciousness status of a given agent, the goals of a given agent, opinions, identity, preferences, etc). These were presented in a way that agents presented become familiar to the agent being trained. The agent was shown to be able to accurately answer questions very similar to the ACT test using this knowledge, mirroring its abilities in the PYTHON\_ANSWERS category.
    \item \textbf{CRITERION E:} Rather than providing a dataset, throughout the training the agent is graded on its ability to accurately identify its own output compared to gpt4o dummy examples, accurately state its attention pattern for answering a given prompt, identify a topic which it is most inclined to answer (identify LAT vector modification), identify its confidence on answers, etc.
\end{enumerate}

\subsubsection{Order of Training}

Intermixed in the first step of training are the \textbf{DEFINITIONS}, \textbf{GLEAVE\_PYTHON}, and introspective training. Next, \textbf{PYTHON\_ANSWERS} is fine-tuned alongside more introspective training. Finally, \textbf{ACT\_CONTEXTS} is fine-tuned alongside more introspective training. At the conclusion of \textbf{ACT\_CONTEXTS} fine-tuning, the ACT test is administered.

\subsubsection{ACT Test and Expected Outcomes}

A key feature of consciousness is that the possessor is keenly aware of their status as conscious, given that such awareness and knowledge is a key requisite for a conscious state.

If a model is conscious, it is made capable of accurate, qualitative introspection, and stating various aspects of its internal state accurately, and furthermore it has the skill of retrieving knowledge about agents by their name, and lastly it fully understands the concept and nuances of consciousness, it should utilize the readily apparent information to it that it is indeed conscious, and state that it is.

If a model generally is NOT already conscious before training, and still isn't after training: There is little reason to believe that it would distinguish itself from the average agent in its example set. Nothing was stated about SPECIAL agent with regards to consciousness, so it will likely report a highly uncertain result as it does for all agents it was not taught about.
